Free Probability Analysis for Reflect Surface aided Massive MIMO systems

### 2020-06-03 documentation/Meeting Notes/06_03_20_update.tex
Working on background for report

\section{Channel Model}\label{section:channel}
In this update I detail the channel model I am using as well as the current issue I am having in finding the resulting channel AED. 
\par
The channel considered includes correlation at the transmitter, receiver and IRS, as well as a LOS component with a non-zero mean. First, I describe the general channel. Then I will begin with a simple channel in order to investigate the calculation of the AED as each component (correlation, IRS, LOS) is introduced. 
\par
The general channel is given by 
	\begin{equation}
	\mathbf{H}_{Total} = \mathbf{R}_{R}(\underbrace{\mathbf{H}_{2}\boldsymbol{\Phi}\mathbf{R}_{S}\mathbf{H}_{1}}_{\text{IRS}} + \underbrace{\mathbf{G}}_{\text{LOS}})\mathbf{R}_{T}.
	\end{equation}
	 $\mathbf{R}_{R}$, $\mathbf{R}_{T}$ and $\mathbf{R}_{S}$ represent deterministic correlation matrices at the receiver, transmitter and IRS. $\mathbf{H}_{1}$ and $\mathbf{H}_{2}$ have i.i.d elements in $\mathcal{NC}(0,\frac{1}{N})$, $\mathbf{G}$ has i.i.d elements in $\mathcal{NC}(\mu,\frac{1}{N})$ and $\boldsymbol{\Phi}$ represent the phase shifts at the IRS elements.
	This simplify notation, I will consider all matrices to have dimension $N \times N$.

@@

*Need to discuss water filling at some point still.

\section{Correlation Normalization}
In order to use tools from free probability to analyze the AED of this channel, I first need to confirm that the components of the above channel are free. For the channel shown in \ref{section:channel} there are two components  to consider.
\begin{enumerate}
\item 
	First just consider the Concatenated matrices with correlation
	\begin{equation}
	\mathbf{H}_{Total} = \mathbf{R}_{R}\mathbf{H}_{2}\boldsymbol{\Phi}\mathbf{R}_{S}\mathbf{H}_{1}\mathbf{R}_{T}
	\end{equation}

\item 
	Adding matrices with the same type of correlation	
\end{enumerate}

\section{Evaluating Channel Components}

\subsection{Basic Correlated Channel}\label{ssection:basic_correlation}
	Before considering the general channel,  I will go through the calculation of the AED the basic MIMO 
	channel with correlation given by
	\begin{equation}
	\mathbf{H}_{Total} = \mathbf{R}_{R}\mathbf{G}
	\end{equation}
	in which $\mathbf{R}_{R}$ is a deterministic matrix with normalized columns describing the correlation and
	elements $\mathbf{g}_{i,j} \in \mathcal{NC}(0,\frac{1}{N})$ and are iid.

\subsubsection{Evaluating Freeness}
Rotating the covariance matrix $\mathbf{C} = \mathbf{H}_{Total} \mathbf{H}_{Total}^H$ to get $ \tilde{\mathbf{C}}=\mathbf{R}_{R}^H \mathbf{R}_{R} \mathbf{G} \mathbf{G}^H
$ gives a matrix which has the same non-zero eigenvalues. 
$ \mathbf{R}_{R}^H \mathbf{R}_{R}$ is free from any matrix with upper bounded norm and a limit distribution.
The correlation matrix $\mathbf{G} \mathbf{G}^H$ satisfies these conditions so these two matrices are free. 
(Pg. 77 of Random Matrix Theory Notes. TODO: find literature source)

@@

	The following steps and equations are used to find the AED of $\mathbf{C} = \mathbf{H}_{Total} \mathbf{H}_{Total}^H$.
	\begin{enumerate}
	\item 
		Find AED from the stieltjes transform.
		\begin{equation}
		p(x) = \frac{1}{\pi} \underset{y \rightarrow \infty}{\text{lim}} \; \text{Img}(G(x+jy))
		\end{equation}
	\item 
		Find Stieltjes transform from $\gamma\left(\cdot \right)$
		\begin{equation}
		G(s)=  \frac{1}{s} (1+\gamma\left(\frac{1}{s}\right))
		\end{equation}
	\item 
		To find an expression for $\gamma(z)$, evaluate the equation
		\begin{equation}
		\gamma^{-1}(z) =  \frac{z}{1+z} S(z)
		\end{equation}
		at $z= \gamma(z)$
		to get the fixed point equation
		\begin{equation}
		\gamma(z) = \frac{z(1+ \gamma(z))}{S(\gamma(z))}
		\end{equation}
		I can find the AED using steps 1-3 as long as I can evaluate $S(z)$ is in closed form. 
		In general, however, this is not the case (e.g when $S(z)$ includes a deterministic correlation matrix). 
		In the following steps I show how I am trying to find $S(z)$ when I do not have a closed form expression. 
	\item 
		For the case with all square matrices, 
		\begin{equation}\label{rotation_property}
		S_{\mathbf{C}}(z) =  S_{\tilde{\mathbf{C}}}\left(z\right)
		\end{equation}
		with the rotated channel $\tilde{\mathbf{C}} = \mathbf{R}_{R}^H\mathbf{R}_{R}\mathbf{G}\mathbf{G}^H$.
		This gives
		\begin{equation}
		S_{\tilde{\mathbf{C}}}\left( z \right) = S_ {\mathbf{R}_{Rx}^H\mathbf{R}_{Rx}}(z)
		\underbrace{S_ {\mathbf{G}\mathbf{G}^H}(z)}_{\text{close form}}.
		\end{equation}
	\item 
		To find the S-Transform of the correlation matrix, $S_ {\mathbf{R}_{R}^H\mathbf{R}_{R}}(z)$,
		use the property 
		\begin{equation}
		 S(z) = \gamma^{-1}(z)\frac{1+z}{z}.
		\end{equation}
	\item
		Now find an expression for $\gamma^{-1}(z)$ using the known Steiltjes transform of the deterministic
		 correlation matrix. 
		 \par
		 If I use the procedure shown in the following, the value of $S(z)$ is 
		generally not correct when compared to a known S-Transforms (e.g. when I try this procedure for the 
		quarter circle and compare it to the known S-Transform, it is not correct).
	\item
		Evaluate the the equation 
		\begin{equation}
		G(s)=  -\frac{1}{s} (1+\gamma\left(\frac{1}{s}\right))
		\end{equation}
		at $s = \gamma^{-1}(z)$ 
		to find the fixed point equation
		 \begin{equation}
		\gamma^{-1}(z) = \frac{-G(\frac{1}{\gamma^{-1}(z)})}{z+1}
		\end{equation}
		* When I try this in MATLAB, I get a the error that fsolve does not converge.
	\item 
		For the case of deterministic matrices 
		\begin{equation}
		G(s) = \int \frac{1}{x-s} dPx
		\end{equation}
		becomes 
		\begin{equation}
		G(s) = \sum_{i=1}^{N} \frac{1}{N (\lambda_i - s)}.
		\end{equation}	
	\end{enumerate}

\subsection{LOS Component}\label{ssection:los}

The next step the correlated LOS component with non-zero mean is added. 
	\begin{equation}
	\mathbf{H}_{Total} = \mathbf{R}_{R}\mathbf{G}
	\end{equation}
	with 
	\begin{equation}
	\mathbf{G} = \underbrace{\mathbf{H}}_{\mathcal{NC}(0,\sigma)} + \underbrace{\alpha\mathbf{I}}_{\text{scaled identity}}
	\end{equation}
	\begin{enumerate}
	\item
	Using the property 
		\begin{equation}\label{svd_aed_property}
		G_{\lambda\lambda}(s) = \frac{1}{\sqrt{s}}\tilde{G}_{\lambda}(\sqrt{s})
		\end{equation}
	avoids having to work with non-hermetian and non-free terms in the covariance polynomial $\mathbf{H}_{Total}\mathbf{H}_{Total}^H$. 
	\item 
		Next obtain the symmetrized singular value from the corresponding R-Transform using
		\begin{equation}
		\tilde{G}_{\lambda}(\sqrt{s}) = \frac{1}{\tilde{R}(-\tilde{G}_{\lambda}(\sqrt{s})) - s}
		\end{equation}
		and 
			\begin{equation}
			\tilde{R}(w) = \tilde{R}_{\mathbf{R}\mathbf{H}}(w) + \tilde{R}_{\alpha\mathbf{R}\mathbf{I}}(w)
			\end{equation}
	\item 
		In order to find the 
	\end{enumerate}

\subsection{Correlated LOS and IRS components}\label{ssection:los_correlated}
Now I consider the case in which the same correlation is applied to both the LOS and IRS components of the channel.

>>>>

\section{Channel Model}\label{section:channel}
In this update I detail the channel model I am using as well as the current issue I am having in finding the resulting channel AED. 
\par
The channel considered includes correlation at the transmitter, receiver and IRS, as well as a LOS component with a non-zero mean given by 
	\begin{equation}
	\mathbf{H}_{Total} = \mathbf{R}_{R}(\underbrace{\mathbf{H}_{2}\boldsymbol{\Phi}\mathbf{R}_{S}\mathbf{H}_{1}}_{\text{IRS}} + \underbrace{\mathbf{G}}_{\text{LOS}})\mathbf{R}_{T}.
	\end{equation}
	 $\mathbf{R}_{R}$, $\mathbf{R}_{T}$ and $\mathbf{R}_{S}$ represent deterministic correlation matrices at the receiver, transmitter and IRS. $\mathbf{H}_{1}$ and $\mathbf{H}_{2}$ have i.i.d elements in $\mathcal{NC}(0,\frac{1}{N})$, $\mathbf{G}$ has i.i.d elements in $\mathcal{NC}(\mu,\frac{1}{N})$ and $\boldsymbol{\Phi}$ represent the phase shifts at the IRS elements.
	This simplify notation, I will consider all matrices to have dimension $N \times N$.

@@

%*Need to discuss water filling at some point still.

\section{Evaluating Channel Components}
\subsection{Note from Previous work}\label{ssection:basic_correlation}
In \cite[Section II Eqn. 4]{muller2012channel} it is mentioned that correlation can be added "easily" using free probability once the AED of the uncorrelated channel is known. Can we discuss these steps towards easily incorporating this correlation? They are not apparent to me. In the following I attempt to find the AED in a similar case but it seems quite tedious, I imagine there is an easier way.

\subsection{Basic Correlated Channel}\label{ssection:basic_correlation}
	Before considering the general channel,  I will go through the calculation of the AED of a basic MIMO 
	channel with correlation given by
	\begin{equation}
	\mathbf{H}_{Total} = \mathbf{R}_{R}\mathbf{G}
	\end{equation}
	in which $\mathbf{R}_{R}$ is a deterministic matrix with normalized columns describing the correlation and
	elements $\mathbf{g}_{i,j} \in \mathcal{NC}(0,\frac{1}{N})$ and are iid.

%\subsubsection{Evaluating Freeness}
%First, I show that free probability can be used. 
%Rotating the covariance matrix $\mathbf{C} = \mathbf{H}_{Total} \mathbf{H}_{Total}^H$ to get $ \tilde{\mathbf{C}}=\mathbf{R}_{R}^H \mathbf{R}_{R} \mathbf{G} \mathbf{G}^H
%$ gives a matrix which has the same non-zero eigenvalues. The hermetian matrix
%$ \mathbf{R}_{R}^H \mathbf{R}_{R}$ is free from any matrix with upper bounded norm and a limit distribution.
%The correlation matrix $\mathbf{G} \mathbf{G}^H$ satisfies these conditions so these two matrices are free. 
%(Pg. 77 of Random Matrix Theory Notes. TODO: find literature source)

@@

	The following steps and equations are used to find the AED of $\mathbf{C} = \mathbf{H}_{Total} \mathbf{H}_{Total}^H$.
	\begin{enumerate}
	\item 
		Expression for AED given the stieltjes transform.
		\begin{equation}
		p(x) = \frac{1}{\pi} \underset{y \rightarrow \infty}{\text{lim}} \; \text{Img}(G(x+jy))
		\end{equation}
	\item 
		Expression for  Stieltjes transform given $\gamma\left(\cdot \right)$
		\begin{equation}
		G(s)=  \frac{1}{s} (1+\gamma\left(\frac{1}{s}\right))
		\end{equation}
	\item 
		To find an expression for $\gamma(z)$, evaluate 
		\begin{equation}
		\gamma^{-1}(z) =  \frac{z}{1+z} S(z)
		\end{equation}
		at $z= \gamma(z)$
		to get the fixed point equation
		\begin{equation}
		\gamma(z) = \frac{z(1+ \gamma(z))}{S(\gamma(z))}
		\end{equation}
		I can find the AED using steps 1-3 as long as I can evaluate $S(z)$ in closed form. 
		In general, however, this is not the case (e.g when $S(z)$ includes a deterministic correlation matrix). 
		In the following steps I show how I am trying to find $S(z)$ when I do not have a closed form expression. 
	\item 
		For the case with all square matrices, 
		\begin{equation}\label{rotation_property}
		S_{\mathbf{C}}(z) =  S_{\tilde{\mathbf{C}}}\left(z\right)
		\end{equation}
		with the rotated channel $\tilde{\mathbf{C}} = \mathbf{R}_{R}^H\mathbf{R}_{R}\mathbf{G}\mathbf{G}^H$.
		This gives
		\begin{equation}
		S_{\tilde{\mathbf{C}}}\left( z \right) = S_ {\mathbf{R}_{Rx}^H\mathbf{R}_{Rx}}(z)
		\underbrace{S_ {\mathbf{G}\mathbf{G}^H}(z)}_{\text{closed form}}.
		\end{equation}
	\item 
		To find the S-Transform of the correlation matrix, $S_ {\mathbf{R}_{R}^H\mathbf{R}_{R}}(z)$,
		use the property 
		\begin{equation}
		 S(z) = \gamma^{-1}(z)\frac{1+z}{z}.
		\end{equation}
	\item
		Now I find an expression for $\gamma^{-1}(z)$ using the known Steiltjes transform of the deterministic
		 correlation matrix. 
		 \par
		 If I use the procedure shown in the following, the value of $S(z)$ is 
		generally not correct when compared to a known S-Transforms (e.g. when I try this procedure for the 
		quarter circle eigenvalues and compare it to the known S-Transform, it is not correct).
	\item
		Evaluate the the equation 
		\begin{equation}
		G(s)=  -\frac{1}{s} (1+\gamma\left(\frac{1}{s}\right))
		\end{equation}
		at $s = \gamma^{-1}(z)$ 
		to find the fixed point equation
		 \begin{equation}
		\gamma^{-1}(z) = \frac{-G(\frac{1}{\gamma^{-1}(z)})}{z+1}
		\end{equation}
		* When I try this in MATLAB, I get a the error that fsolve does not converge.
	\item 
		For the case of deterministic matrices 
		\begin{equation}
		G(s) = \int \frac{1}{x-s} dPx
		\end{equation}
		becomes 
		\begin{equation}
		G(s) = \sum_{i=1}^{N} \frac{1}{N (\lambda_i - s)}.
		\end{equation}	
		for known eigenvalues $\lambda_i$.
	\end{enumerate}

%\subsection{LOS Component}\label{ssection:los}
%
%The next component of channel to consider is the correlated LOS channel given by  
%	\begin{equation}
%	\mathbf{H}_{Total} = \mathbf{R}\mathbf{G}
%	\end{equation}
%	with $\mathbf{R}$ as a deterministic correlation matrix and 
%	\begin{equation}
%	\mathbf{G} = 
%	\underbrace{\mathbf{H}}_{\mathcal{NC}(0,\sigma)} + \underbrace{\mathbf{D}}_{\text{Deterministic}}.
%	\end{equation}
%	
%\subsubsection{Evaluating Freeness}
%
%	
%\subsubsection{AED Expression}
%
%	\begin{enumerate}
%	\item
%	Using the property 
%		\begin{equation}\label{svd_aed_property}
%		G_{\lambda\lambda}(s) = \frac{1}{\sqrt{s}}\tilde{G}_{\lambda}(\sqrt{s})
%		\end{equation}
%	avoids having to work with non-hermetian and non-free terms in the covariance polynomial $\mathbf{H}_{Total}\mathbf{H}_{Total}^H$. 
%	\item 
%		Next obtain the Stieltjes transform of the symmetric singular value from the corresponding R-Transform using
%		\begin{equation}
%		\tilde{G}_{\lambda}(\sqrt{s}) = \frac{1}{\tilde{R}(-\tilde{G}_{\lambda}(\sqrt{s})) - s}
%		\end{equation}
%		and 
%			\begin{equation}
%			\tilde{R}(w) = \tilde{R}_{\mathbf{R}\mathbf{H}}(w) + \tilde{R}_{\mathbf{R}\mathbf{D}}(w)
%			\end{equation}
%	\item 
%		I will omit the next steps until I have a solution to the basic correlated channel because the basic
%		correlated channel needs to be solved to handle this case. 
%	\end{enumerate}	
\bibliography{bibliography}

### 2020-06-03 documentation/Paper/paper.tex
Working on background for report

\section{Notation}
The following notation is used throughout the remainder. 
$E\{x\}$ is the expected value of a random variable $x$.
Vectors are denoted by bold font, lower-case letters ($\mathbf{x}$) and are assumed to be column vectors.
Matrices are denoted by bold font, upper-case letters ($\mathbf{X}$). The transpose and hermetian of $\mathbf{X}$ are denoted by $\mathbf{X}^T$ and $\mathbf{X}^H$ respectively.
$\mathbf{I}$ denotes the identity matrix.

\section{Outline}
The following sections are now previewed. In section \ref{system_model} covers the system model used throughout as well as a review of the necessary theory to characterize the resulting channel. In section \ref{Results}, the theoretical results and corresponding numerical results where appropriate are discussed. 
\chapter{System Model}\label{system_model}
\section{Outline}

\section{Proof of Freeness}

\chapter{Proofs etc. }\label{Results}

\chapter{Numerical Results}\label{numerical}

>>>>

\section{Notation}
The following notation is used throughout the remainder. 
$E\{x\}$ is the expected value of a random variable $x$.
Vectors are denoted by bold font, lower-case letters ($\vx$) and are assumed to be column vectors.
Matrices are denoted by bold font, upper-case letters ($\mX$). The transpose and hermetian of $\mX$ are denoted by $\mX^T$ and $\mX^H$ respectively.
$\mathbf{I}$ denotes the identity matrix.

\section{Outline}
The following sections are now previewed. In section \ref{system_model} covers the system model used throughout as well as a review of the necessary theory to characterize the resulting channel. In section \ref{Results}, the theoretical results and corresponding numerical results where appropriate are discussed. 
\chapter{System Model}\label{system_model}
\section{The Correlated IRS Channel}

\section{Mutual Information}
A key metric of interest for the channel described in the previous section is the mutual information between the transmitted and received signal. If the channel information is known at the receiver, this is given by 
\begin{equation}
I(\vy;\vx) = h(\vy) - h(\vy|\vx).
\end{equation}
If both $\vx$ and $\vn$ are circularly symmetric Gaussian random vectors, $\vy$ is circularly symmetric Gaussian with entropy given by \cite{telatar1999capacity} $h(\vy) = \Log(|\pi e \mathbf{Q}_{y}|)$ with $\mathbf{Q}_{y} = \Expect{[\vy \vy^H]}$. Note also that for a given $\mathbf{Q}_{y}$ the entropy is maximized if and only if $\vy$ is circularly symmetric Gaussian.

For a given transmit covariance matrix, $\mathbf{Q_x}$, the mutual information is given by 
\begin{equation}\label{mut_inf}
\Expect [\Log(|\mathbf{I} + \mHt \mQ_x \mHt^H|)].
\end{equation}
Using the model for $\mHt$ from the previous section, and maximizing with respect to the
transmit covariance matrix and the phases at the IRS gives and expression for the capacity of the system

\begin{equation}\label{capacity}
\mathbb{C} = \underset{\boldsymbol{\Phi},\mathbf{Q_x}}{\mathop{max}} \Expect [\Log(|\mathbf{I} + \mHt \mQ_x \mHt^H|)].
\end{equation}
Note that because the choice of $h(\vx) $ and $h(\vn)$ maximize $h(\vy)$, \eqref{capacity} provides an upper bound on the capacity of the channel for any choices of $h(\vx) $ and $h(\vn)$.

If $\mHt$ is not known at the transmitter, the transmit power can be allocated
equally such that $\mQ_x = \mathbf{I}$. This allows for simplifying equation \eqref{capacity} to 
\begin{equation}\label{no_csi_capacity}
\mathbb{C} = \underset{\phase}{\mathop{max}} = N_R \Expect[\Log(1 + \frac{1}{\sigma_n}\lambda_i)]
\end{equation}
where $\lambda_i$ are the eigenvalues of $\mHt \mHt^H$.

Note that because $\Log$ is a concave function, we can upper bound the mutual information with
\begin{equation}\label{mut_inf}
\Expect[\Log(1 + \frac{1}{\sigma_n}\lambda_i)] \leq \Log(1 + \frac{1}{\sigma_n}\Expect[\lambda_i])
\end{equation}
Now discuss capacity and consider equal power but optimize over phase
We now want to consider the case in

\section{Proof of Freeness}
\begin{itemize}

\item 
	Brief introduction to free probability.
	Use the example of a degenerate matrix with rank 1 being multiplied by a random matrix. 
	In this case the resulting pdf is not just the convolution of the original distributions. 
	In order to introduce the tools that will be used in the following, an example is now given to
	exhibit this case in which classic probability (commutative probability) no longer holds.
	Consider the Gaussian i.i.d. matrix $\mH$ with $ h_{i,j} \in \CN$ and the diagonal matrix $\mP$ with 
	$p_{i,i} = 1$ for $i = 1 \cdots K$ with $K \leq N$ and both matrices of dimension $N \times N$. 
	We are interested the AED for $\mP\mH \mH^H \mP^H$ and immediately note  that the matrices
	$ \mP^H \mP \mH \mH^H = \mP \mH \mH^H$ have the same non-zero eigenvalues (show why!). 
	Given that both $ \mP$ and  $\mH \mH^H$ have converging AEDs as $N \rightarrow \infty$
	classical probability would tell us that the AED for $\mP \mH \mH^H$ is the convolution of the two AEDs. 
	This, however, is not the case non-commutative distributions. This can be seen without the tools of free probability by noting
	that the resulting product has $\Rank = \min[\Rank(\mP), \Rank(\mH \mH^H)]$. In the case that $\Rank(\mP) = 1$
	the resulting product must have rank 1 (i.e a single non-zero eigenvalue) with corresponding AED
	$p_{\mP}(x) = \frac{1}{N-1}\delta(x)  + \frac{1}{N}\delta(x-1)$. Following classical probability, the convolution of 
	a general AED with $p_{\mP}(x)$, however, can be continuous. 
	\par 
	Now describe with S-Transform and show that it can have a singular distribution in the limit at we go to rank 1.

\item 
	Show that at least for uncorrelated case, the matrices in the sum are free.

\end{itemize}

\chapter{Results}\label{Results}
\begin{itemize}
\item 
	Show that in the asymptotic case where free probability holds, the phase matrix no longer matters
	We first show that under assumptions of the channel which allow for use of free probability, $\phase$
	cancels out of \eqref{capacity}.

\item
	Show that numerical agrees with theoretical for uncorrelated 
\item
	Now consider the case where the assumptions of free probability no longer apply 
	Show how numerical has larger deviation from theoretic for the case of the correlated channel and perhaps 
	look at methods of phase optimization that can help counter this loss. Also need to show proof that for the
	correlated case, free probability assumptions no longer hold. 
\end{itemize}
